Forecasting the 2024 U.S. Presidential Election: Kamala Harris' Projected Victory

### 2024-11-04 paper/paper.qmd
committing text changes to paper.qmd in appendix

# Data {#sec-data}
## Overview
The data for this paper contains the polling information of the 2024 Presidential Election cycle and was obtained from Project 538 [@FiveThirtyEight] on October 19th, 2024. The data is periodically updated throughout the presidential election campaign to reflect current and most up-to-date polling results. It consists of polling information from various pollsters such as the New York Times, ActiVote, Morning Consult, etc. For every poll conducted, relevant variables such as **'state'**, **'start_date'**, **'end_date'**, **'transparency_score'**, **'candidate_name'**, and percentage of support (**'pct'**) for the appropriate candidates are included. The data compiles information from all major pollsters in the US and provides us with an understanding of the current state of the presidential election.

The present analysis focuses on Donald Trump and Kamala Harris, the two leading candidates in the current presidential election race. For all the variables mentioned (see @sec-variables), we collected observations for both of the candidates. We collect observations from the data where the **‘candidate_name’** is Trump and Harris. We are interested in the support for both candidates over time, for each state, therefore, we collect observations of **‘end_date’** and **‘state’**, as well as the electorate support for each candidate, which is denoted as **‘pct’**. Furthermore, since there are many pollsters present in the data, we extracted pollsters with high-quality data that is reliable, which is denoted as having a high **‘numeric_grade’** and **‘transparency_score’**, details of data cleaning are further explained in [Appendix -@sec-data-details].

@@

## Measurement
Voter opinions are influenced by various factors, such as media, political events, personal beliefs, and social pressures, which can change frequently, especially near an election. Pollsters simplify these complex opinions into basic polling responses like “Trump,” “Harris,” or “Undecided,” which can obscure nuanced voter sentiment. This initial conversion may not capture voters’ internal conflicts or reservations, as respondents must choose a definitive option.

These individual responses are aggregated into percentages, treating each “Trump” or “Harris” response equally, regardless of confidence level, further smoothing out complexities. The model uses these aggregated support percentages to predict election outcomes by estimating the likelihood of each candidate winning. This step assumes that poll data accurately reflects voter behavior on Election Day, which may not always hold true, as it simplifies support into binary outcomes (predicting Trump or Harris as the winner based on higher support).

@@

To give an overview of the outcome variable, divided per candidate, we provide summary statistics of this key variable, highlighting the characteristics of the polling support for each candidate across the battleground states. 
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: tbl-outcomevar
#| tbl-cap: "Summary Statistics for Trump and Harris Polling Data - Presents key summary statistics, including total polls, average support percentages, and the maximum and minimum polling levels for both candidates across the specified battleground states. All values are rounded to three decimal places, providing an overview of the polling landscape as of the selected election date."

@@

## Model Setup
Modeling processes are conducted by employing the following packages: `rstanarm` package of @rstanarm, `brms` package of  @brms for Bayesian regression modeling, `Matrix` package of @Matrix for handling sparse and dense matrices, `splines` package of @splines for regression splines, `caret` package of @caret for machine learning support, and `broom.mixed` package of @broom.mixed for tidying mixed models.

@@

## Sample Recruitment Methods
Each poll is conducted by phone using live interviewers at call centers based in Florida, New York, South Carolina, Texas, and Virginia. Most people that the live interviewers call do not answer the call, and some don’t finish the survey, leaving only about two percent both answer their phone and complete the poll. To mitigate this:
First, the pollster keeps calling numbers until someone picks up, even though it can take hours–a reason why a high-quality poll is time-consuming and expensive to conduct. 
Second, voter files are used to help make sure the pollster is reaching a representative sample; for example, they may reach out to a higher percentage of people who are less likely to respond. 
Third, pollster mathematically adjusts the responses they do receive so they are representative of the wider population, it includes adjusting by age, race, gender, voting history and etc.

>>>>

# Data {#sec-data}
## Overview
The data for this paper, obtained from Project 538 [@FiveThirtyEight] on October 19, 2024, encompasses polling information for the 2024 Presidential Election cycle, reflecting the most current polling results from various pollsters, including the New York Times, ActiVote, and Morning Consult. Key variables include **'state'**, **'start_date'**, **'end_date'**, **'transparency_score'**, **'candidate_name'**, and 'percentage of support (**'pct'**) for each candidate. Our analysis focuses on Donald Trump and Kamala Harris, the leading candidates, collecting observations where ‘candidate_name’ matches either Trump or Harris. We aim to track support over time for each candidate by extracting relevant data on **‘end_date’**, **‘state’**, and **‘pct’**. To ensure reliability, we included only polls from pollsters with high **‘numeric_grade’** and **‘transparency_score’**. Details on the data cleaning process are provided in [Appendix -@sec-data-details].

@@

## Measurement
Voter opinions are influenced by various factors, such as media, political events, personal beliefs, and social pressures, which can change frequently, especially near an election. Pollsters simplify these complex opinions into basic polling responses like “Trump,” “Harris,” or “Undecided,” which can obscure nuanced voter sentiment. This initial conversion may not capture voters’ internal conflicts or reservations, as respondents must choose a definitive option. These individual responses are aggregated into percentages, treating each “Trump” or “Harris” response equally, regardless of confidence level, further smoothing out complexities. The model uses these aggregated support percentages to predict election outcomes by estimating the likelihood of each candidate winning. This step assumes that poll data accurately reflects voter behavior on Election Day, which may not always hold true, as it simplifies support into binary outcomes (predicting Trump or Harris as the winner based on higher support).

@@

To give an overview of the outcome variable, divided per candidate, we provide summary statistics of this key variable, highlighting the characteristics of the polling support for each candidate across the battleground states. 
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: tbl-outcomevar
#| tbl-cap: "Summary statistics of polling data, showing a total of 107 polls conducted. Average support is 47.801% for Donald Trump and 47.999% for Kamala Harris, with similar maximum and minimum support levels, indicating a competitive electoral landscape."

@@

## Model Setup
Modeling processes are conducted by employing the following packages: `rstanarm` package of @rstanarm, `brms` package of  @brms for Bayesian regression modeling, `Matrix` package of @Matrix for handling sparse and dense matrices, `splines` package of @splines for regression splines, `caret` package of @caret for machine learning support, and `broom.mixed` package of @broommixed for tidying mixed models.

@@

## Sample Recruitment Methods
We will initiate our survey by sending invitations to the 2,000 selected voters via email. To enhance response rates and reduce non-response bias, we will send a reminder email three days later. If we do not receive a response within a week (7 days), we will follow up by phone to reach those who did not respond to the email invitation. To effectively manage the recruitment process, we will allocate `$100,000` of the budget for staffing. We will employ a small team of around 30 people responsible for sending invitations during the first three days and following up with non-respondents over the entire two-week process. Each employee will be paid `$24` per hour for 40 hours per week for two weeks, resulting in a total of 40 hours worked per employee and `$960` for each, `$28,800` in total. Additionally, we will allocate `$86,000` for the phone follow-up process as well as the data cleaning and analysis process. We will employ a team of 20 data scientists to perform data engineering and analysis, with them being paid $3560 each for the duration of their work term, which will be around three to five weeks.

Overall, this recruitment method is designed to be both efficient and effective. By reaching out to a diverse group of voters and encouraging participation through personalized follow-ups, we can collect high-quality data that accurately reflects the opinions and behaviors of the broader population. Additionally, this recruitment method minimizes early-stage labor costs by using email as the initial contact method, reserving more follow-up efforts for those who have not yet responded after a week. This approach not only helps a higher participation rate but also improves the reliability of the data collected.
